Outline of Ideas for BTC Transaction Frequency

### 2015-10-13 outline/outline.tex
Fixed some of the language and notation describing how a block arrives and what happens when it arrives

\section{Motivation}
As Bitcoin becomes more popular, the transaction rate increases. Since existing Bitcoin rules limit the amount of transactions per block to one MB at some point the blocks will begin filling up. Bitcoin blocks are by design distributed according a Poisson process with a rate of discovery of one block every 10 minutes. Let $\tau$ index time in minutes, and $N(\tau)$ denote the number of blocks discovered by time $\tau$. Then for $\delta > 0$, the number of blocks discovered between time $\tau$ and time $\tau + \delta$ is Poisson distributed, that is $N(\tau + \delta) - N(\tau) \sim Poi(\delta/10)$. Let $k=1,2,\dots$ denote each of the blocks, in order. Then the amount of time between blocks is exponentially distributed. Let $\tau_k$ denote the the time when block $k$ arrives. Then $\delta_k \equiv \tau_{k+1} - \tau_{k} \stackrel{iid}{\sim} Exp(1/10)$ for $k=1,2,\dots,K$, where if $x\sim Exp(\lambda)$ it has the density $p(x) = \lambda e^{-x\lambda}$ for $x>0$ and $\lambda>0$ with mean $1/\lambda$.

There is always some chance that a given block will fill up since arrival times are stochastic. Specifically, for the $k$'th block this probability is
\[
P\left(x_{\tau_k + \delta_k} - x_{\tau_k} > 1\right).
\]
This probability depends crucially on the process that $x_\tau$ follows. If we assume that $x_{\tau}$ is determinstic with a constant transaction rate of $\gamma$ MB/min, then 
\[
x_{\tau_k + \delta_k} - x_{\tau_k} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/\gamma 10)
\]
and the probability of a full block is available in closed form as
\[
P\left(x_{\tau_k + \delta_k} - x_{\tau_k} > 1\right) = \int_1^\infty 1/10\gamma e^{-1/10\gamma x}dx = e^{-1/10\gamma}.
\]

Clearly a constant transaction rate does not capture Bitcoin's behavior since it has been growing in popularity. With a non-constant and potentially stochastic transaction rate, the probability of a block filling up is no longer necessarily available in closed form. Using a Monte Carlo simulation of $N$ draws of the block sizes from the process $x_{\tau}$, we can approximate the probability of block $k$ filling up with
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{x^{(i)}_{\tau_{k}^{(i)}+\delta_k^{(i)}} - x^{(i)}_{\tau_k^{(i)}} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function. If we let $\gamma$ denote the mean transaction rate, which is the same as the actual transaction rate in the constant, deterministic case, we can alter $\gamma$ to see how the probability of a full block changes under a variety of regimes.

A key question is whether the assumption of a constant transaction rate is approximately correct on a short enough time-scale. In order to test this assumption we fit a the gamma distribution to the block sizes of various pieces of the block chain. We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(\alpha,\beta)$, it has the density
\[
p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
\]
for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$. Since the $G(1,\beta)$ and $Exp(\beta)$ distributions are identical, if we find that $\alpha\neq 1$ then transaction rates cannot be constant. We focus on the most recent 5000 blocks in our database, which is in the process of downloading the entire chain, ending with block 303151.  For priors we will assume $\log \alpha \sim N(0, 10)$ and $\beta \sim G(1, 1)$. This prior on $\alpha$ implies that the prior median is $\alpha=1$ so that the null hypothesis is favored relatively strongly by the prior. We fit the model to all 5000 blocks, the first 1000 blocks, the second 1000 blocks, etc., and the first 500 blocks, the second 500 blocks, etc. The results are in Table \ref{tab:gammafit}.

@@

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block is completed. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time it is the current block so that $\tau_{k+1}=\tau_k + \delta_k$, and now $t_k$ the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k} + \delta_k} - x_{\tau_k}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing, infinitely divisible\footnote{Should we define?} stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider the classes of non-negative infinite activity\footnote{Should we define?} L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.\footnote{Cite the Levy process chapter here}. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.

@@

It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error. This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_k + \delta_k}$ without resorting to approximation methods. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments. We will use the gamma process since it is widely available in statistical packages, unlike the main alternative, the inverse Gaussian.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar model that avoids this problem is the lognormal distribution, which we will use.

Combining these pieces we obtain the following model. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k+1}} - x_{\tau_{k}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
\delta_{k} = \tau_{k+1} - \tau_{k} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. Note that $x_{\tau_{k+1}} - x_{\tau_k}$ depends on $\delta_k$ because the $k$'th block arrives at $\tau_k$ and departs at $\tau_{k+1}=\tau_k + \delta_k$ so that block $k$ has $\delta_k$ minutes to accumulate transaction data. To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$. Table \ref{tab:modelfit} contains the results of fitting the model.

@@

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases. We can mitigate 1) while making 2) if not worse, different. Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution.\footnote{Cite Gelman et al. text here} In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ by some amount $\psi$ so that the posterior mean of the shifted distribution of $\gamma$ is the desired amount. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different. We estimate the probability of a block filling using all three approaches, but because of the required assumptions the estimates become less credible for values of $\gamma$ or $\bar{\gamma}$ further away from the estimated values.

>>>>

\section{Motivation}
As Bitcoin becomes more popular, the transaction rate increases. Since existing Bitcoin rules limit the amount of transactions per block to one MB at some point the blocks will begin filling up. Bitcoin blocks are by design distributed according a Poisson process with a rate of discovery of one block every 10 minutes. Let $\tau$ index time in minutes, and $N(\tau)$ denote the number of blocks discovered by time $\tau$. Then for $\delta > 0$, the number of blocks discovered between time $\tau$ and time $\tau + \delta$ is Poisson distributed, that is $N(\tau + \delta) - N(\tau) \sim Poi(\delta/10)$. Let $k=1,2,\dots$ denote each of the blocks, in order. Then the amount of time between blocks is exponentially distributed. Let $\tau_k$ denote the the time when block $k$ arrives. Then $\delta_k \equiv \tau_{k} - \tau_{k-1} \stackrel{iid}{\sim} Exp(1/10)$ for $k=1,2,\dots,K$, where if $x\sim Exp(\lambda)$ it has the density $p(x) = \lambda e^{-x\lambda}$ for $x>0$ and $\lambda>0$ with mean $1/\lambda$. The longer $\delta_k$ the more time for transactions to add up before being added to the block chain as part of block $k$.

There is always some chance that a given block will fill up since arrival times are stochastic. Specifically, for the $k$'th block this probability is
\[
P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right).
\]
This probability depends crucially on the process that $x_\tau$ follows. If we assume that $x_{\tau}$ is determinstic with a constant transaction rate of $\gamma$ MB/min, then 
\[
x_{\tau_k} - x_{\tau_{k-1}} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/\gamma 10)
\]
and the probability of a full block is available in closed form as
\[
P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right) = \int_1^\infty 1/10\gamma e^{-1/10\gamma x}dx = e^{-1/10\gamma}.
\]

Clearly a constant transaction rate does not capture Bitcoin's behavior since it has been growing in popularity. With a non-constant and potentially stochastic transaction rate, the probability of a block filling up is no longer necessarily available in closed form. Using a Monte Carlo simulation of $N$ draws of the block sizes from the process $x_{\tau}$, we can approximate the probability of block $k$ filling up with
\[
\frac{1}{N}\sum_{i=1}^N\mathbbm{1}\left\{x^{(i)}_{\tau_{k}^{(i)}} - x^{(i)}_{\tau_{k-1}^{(i)}} > 1\right\}
\]
where $i$ indexes the Monte Carlo simulations of each of the random quantities inside the indicator function. If we let $\gamma$ denote the mean transaction rate, which is the same as the actual transaction rate in the constant, deterministic case, we can alter $\gamma$ to see how the probability of a full block changes under a variety of regimes.

A key question is whether the assumption of a constant transaction rate is approximately correct on a short enough time-scale. In order to test this assumption we fit a gamma distribution to the block sizes of various pieces of the block chain. We use the shape-rate parameterization of the gamma distribution so that if $x\sim G(\alpha,\beta)$, it has the density
\[
p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-x\beta}
\]
for $x>0$ with mean $\alpha/\beta$ and variance $\alpha/\beta^2$. Since the $G(1,\beta)$ and $Exp(\beta)$ distributions are identical, if we find that $\alpha\neq 1$ then transaction rates cannot be constant. We focus on the most recent 5000 blocks in our database ending with block 303151 --- we are in the process of downloading the entire chain.  For priors we will assume $\log \alpha \sim N(0, 10)$ and $\beta \sim G(1, 1)$. This prior on $\alpha$ implies that the prior median is $\alpha=1$ so that the null hypothesis is favored relatively strongly by the prior. We fit the model to all 5000 blocks, the first 1000 blocks, the second 1000 blocks, etc., and the first 500 blocks, the second 500 blocks, etc. The results are in Table \ref{tab:gammafit}.

@@

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block arrives. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing, infinitely divisible\footnote{Should we define?} stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider the classes of non-negative infinite activity\footnote{Should we define?} L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$.\footnote{Cite the Levy process chapter here}. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the Gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$.

@@

It is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form since the observation times are observed with error. This simplifies Markov chain Monte Carlo methods that may be used to fit the model by allowing us to exactly simulate $x_{\tau_{k-1} + \delta_k}$ without resorting to approximation methods. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the parameter increments. We will use the gamma process since it is widely available in statistical packages, unlike the main alternative, the inverse Gaussian.

On the data level we have measurement error with a lower bound, motivating a class of models of the form 
\[
t_{k+1} |t_{1:k}, \tau_{1:K} \stackrel{ind}{\sim} p(t|\tau_{k+1}, \phi)\mathbbm{1}\left\{t > m_{k+1})\right\}
\]
where $p(t|\tau,\phi)$ is a class of probability densities on $t>0$ with centrality parameter $\tau$ and dispersion parameter $\phi$, and $m_k = \mathrm{median}(t_{k - 0:10})$. For example $\tau$ could be the mean or the mode and $\phi$ the variance or precision. This interpretation of these parameter applies for the {\it untruncated} distribution $p(t|\tau,\phi)$, but not necessarily for the condtional dsitribution of $t_k$. A natural choice here is another gamma distribution, since the untruncated distribution still must be non-negative. However the gamma causes numerical problems at times since it forces the latent measurement times, $\tau_k$ to enter the gamma function. A similar model that avoids numerical problems is the lognormal distribution, which we will use.

Combining these pieces we obtain the following model. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k-1} + \delta_k} - x_{\tau_{k-1}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
\delta_{k} = \tau_{k} - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$. Table \ref{tab:modelfit} contains the results of fitting the model.

@@

This approach has two major shortfalls: 1) it ignores parameter uncertainty, and 2) it assumes that the other parameters will remain constant as $\gamma$ increases. We can mitigate 1) while making 2) if not worse, different. Specifically, given access to an approximate simulation of the posterior of the model parameters, we can simulate the model for each draw from the posterior in order to estimate the probability of a block filling up --- this is the posterior predictive distribution.\footnote{Cite Gelman et al. text here} In order to allow for different transaction rates in this simulation, we can set $\gamma$ at a particular value while using the full posterior of the other model parameters, or even better we can shift the posterior of $\gamma$ so that the posterior mean of the shifted posterior is a desired value. This approach takes into account parameter uncertainty but assumes that our uncertainty about the other parameters would be unchanged if the location of $\gamma$'s distribution were different. We estimate the probability of a block filling using all three approaches, but because of the required assumptions the estimates become less credible for values of $\gamma$ or $\mathrm{E}[\gamma]$ further away from the estimated values.
